Types from data: \textnormal{Making structured data first-class citizens in F\#}

### 2015-11-20 pldi-2016/paper.tex
Final changes in section 6

In this section, we discuss our practical experience with the safety guarantees provided by the
F\# Data type providers and other notable aspects of the implementation.

@@

The \emph{relativized safety} property does not guarantee the same amount of safety as traditional
ML type safety, but it reflects the reality of programming with external data sources that is 
increasingly important \cite{age-of-web}. Type providers do not reduce the safety -- they simply 
illuminate the existing issues.

One such issue is the handling of schema change. With type providers, the sample is captured 
at compile-time. If the schema changes later (so that the input is no longer a subtype of the 
sample), the program fails at runtime and developers have to handle the exception. This is the 
same problem that happens when reading data using any other library.

@@

In our formalization, we treat XML, JSON and CSV uniformly as \emph{structural values}. With the addition of 
named records (for XML nodes), the definition of structural values is rich enough to capture all 
three formats\footnote{The same mechanism has later been used by the HTML type provider 
(\url{http://fsharp.github.io/FSharp.Data/HtmlProvider.html}), which provides similarly easy 
access to data in HTML tables and lists.}. However, parsing real-world data poses a number of practical issues.

\paragraph{Reading CSV data.}
When reading CSV data, we read each row as an unnamed record and return a collection of rows.
One difference between JSON and CSV is that in CSV, the literals have no data types and so 
we also need to infer the type of primitive values. For example:
%
{\small{
\begin{verbatim}
  Ozone, Temp, Date,       Autofilled
  41,    67,   2012-05-01, 0
  36.3,  72,   2012-05-02, 1
  12.1,  74,   3 kveten,   0
  17.5,  #N/A, 2012-05-04, 0
\end{verbatim}
}}
%
\noindent
The value {\small\ttfamily \#N/A} is commonly used to represent missing values in CSV and is treated
as \kvd{null}. The \ident{Date} column uses mixed formats and is inferred as \ident{string} 
(we support many date formats and ``May 3'' would be parsed correctly). More interestingly,
we also infer \ident{Autofiled} as Boolean, because the sample contains only $0$ and $1$.
This is handled by adding a \ident{bit} type which is a subtype of both \ident{int} and \ident{bool}.

\paragraph{Reading XML documents.}
Mapping XML documents to structural values is perhaps the most interesting. For each node, we
create a named record. Attributes become record fields and the body becomes a field with a special
name:
%
{\small{
\begin{verbatim}
  <root id="1">
    <item>Hello!</item>
  </root>    
\end{verbatim}
}}
%
\noindent
This XML becomes a record \ident{root} with fields \ident{id} and $\bullet$ for the body. 
The nested element contains only the $\bullet$ field with the inner text. As with CSV, we
infer type of primitive values:
%
\begin{equation*}
\ident{root}~\{ \ident{id} \mapsto 1, \bullet \mapsto [ \ident{item}~\{ \bullet \mapsto \str{Hello!} \}] \}
\end{equation*}
%
When generating types for XML documents, we also lifts the members nested under the $\bullet$ 
field into the parent type to simplify the structure of the generated type.

The XML type provider also includes an option to use \emph{global inference}. In that case, 
the inference from values (\ref{sec:formal-inferval}) unifies the types of all records with the 
same name. This is useful because, for example, in XHTML all {\small\ttfamily <table>} elements
will be treated as values of the same type.

@@

When introducing type providers (\S\ref{sec:providers-sum}), we mentioned that F\# Data 
implements a special handling of heterogeneous collections. This allows us to avoid inferring a 
variant types in many common scenarios. In the earlier example, a sample collection 
contains a record (with \strf{pages} and \strf{page} fields) and a nested collection with values.

Rather than storing a single type for the collection elements as in $[\sigma]$, heterogeneous
collections store multiple possible element types together with their \emph{  inferred multiplicity} 
(zero or one, exactly one, zero or more):
%
\begin{equation*}
\begin{array}{rcl}
 \psi &\narrow{=}& 1? \lsep 1 \lsep \ast \\
 \sigma &\narrow{=}& ~\ldots \lsep [\sigma_1, \psi_1 | \ldots | \sigma_n, \psi_n ]
\end{array}
\end{equation*}
%
We omit the details, but finding a common supertype of two heterogeneous 
collections is analogous to the handling of variants. We merge cases with the same tag (by finding 
their common super-type) and calculate their new shared multiplicity (for example, by turning 
$1$ and $1?$ into $1?$).

@@

\paragraph{Extending the type systems.} 
Previous work that integrates external data, such as XML \cite{xduce,xduce-ml} and databases 
\cite{links,linq}, into a programming language, required the user to define the schema
or it has an ad-hoc extension that reads the schema.

C$\omega$ \cite{comega-xs} is the most similar to F\# Data. It extends C\# with types similar to our
structural types (including similar heterogeneous collections), but it does not infer the types
from a sample and extends the type system of the host language (rather than using general purpose
embedding).

@@

The PADS project \cite{pads-dsl,pads-ml} tackles a more general problem of handling \emph{any} data format.
The schema definitions in PADS are similar to our structural type. The structure inference for LearnPADS
\cite{pads-learn} infers the data format from a flat input stream. A PADS type provider could follow
many of the patterns we explore in this paper, but formally specifying the safety property would be
challenging.

@@

We explored the F\# Data type providers for XML, CSV and JSON. As most real-world data do not come 
with an explicit schema, the library uses \emph{type inference} that deduces a type from a set of 
samples. Our type inference algorithm is based on a common supertype relation. For usability 
reasons, it prefers records and avoids variant types. The algorithm is predictable, which is 
important as developers need to understand how changing the samples affects the resulting types.

We explore the programming language theory behind type providers. F\# Data is a prime example of 
type providers, but our work also demonstrates a more general point. The types generated by type 
providers can depend on external input and so we can only guarantee \emph{relativized safety}, 
which says that a program is safe only if the actual inputs satisfy additional conditions --
in our case, they have to be subtypes of one of the samples.

Type providers have been described before, but this paper is novel in that it explores concrete 
type providers from the perspective of programming language theory. This is particularly important
for F\# Data type providers, which have been widely adopted by the industry.

>>>>

In this section, we discuss our experience with the safety guarantees provided by the
F\# Data type providers and other notable aspects of the implementation.

@@

The \emph{relativized safety} property does not guarantee the same amount of safety as traditional
closed-world type safety, but it reflects the reality of programming with external data that is 
increasingly important \cite{age-of-web}. Type providers do not reduce the safety -- they simply 
illuminate the existing issues.

One such issue is the handling of schema change. With type providers, the sample is captured 
at compile-time. If the schema changes later (so that the input is no longer related to the  
shape of the sample), the program fails at runtime and developers have to handle the exception. 
The same problem happens when using weakly-typed techniques.

@@

In our formalization, we treat XML, JSON and CSV uniformly as \emph{data values}. With the addition of 
names for records (for XML nodes), the definition of structural values is rich enough to capture all 
three formats\footnote{The same mechanism has later been used by the HTML type provider 
(\url{http://fsharp.github.io/FSharp.Data/HtmlProvider.html}), which provides similarly easy 
access to data in HTML tables and lists.}. However, parsing real-world data poses a number of practical issues.

\paragraph{Reading CSV data.}
When reading CSV data, we read each row as an unnamed record and return a collection of rows.
One difference between JSON and CSV is that in CSV, the literals have no data types and so 
we also need to infer the shape of primitive values. For example:
%
{\small{
\begin{verbatim}
  Ozone, Temp, Date,       Autofilled
  41,    67,   2012-05-01, 0
  36.3,  72,   2012-05-02, 1
  12.1,  74,   3 kveten,   0
  17.5,  #N/A, 2012-05-04, 0
\end{verbatim}
}}
%
\noindent
The value {\small\ttfamily \#N/A} is commonly used to represent missing values in CSV and is treated
as \kvd{null}. The \ident{Date} column uses mixed formats and is inferred as \ident{string} 
(we support many date formats and ``May 3'' would be parsed as date). More interestingly,
we also infer \ident{Autofiled} as Boolean, because the sample contains only $0$ and $1$.
This is handled by adding a \ident{bit} shape which is preferred of both \ident{int} and \ident{bool}.

\paragraph{Reading XML documents.}
Mapping XML documents to structural values is perhaps the most interesting. For each node, we
create a record. Attributes become record fields and the body becomes a field with a special
name:
%
{\small{
\begin{verbatim}
  <root id="1">
    <item>Hello!</item>
  </root>    
\end{verbatim}
}}
%
\noindent
This XML becomes a record \ident{root} with fields \ident{id} and $\bullet$ for the body. 
The nested element contains only the $\bullet$ field with the inner text. As with CSV, we
infer shape of primitive values:
%
\begin{equation*}
\ident{root}~\{ \ident{id} \mapsto 1, \bullet \mapsto [ \ident{item}~\{ \bullet \mapsto \str{Hello!} \}] \}
\end{equation*}
%
When generating F\# types for XML documents, we also lifts the members nested under the $\bullet$ 
field into the parent type to simplify the structure of the generated type.

The XML type provider also includes an option to use \emph{global inference}. In that case, 
the inference from values (\S\ref{sec:formal-inferval}) unifies the shapes of \emph{all} records with the 
same name. This is useful because, for example, in XHTML all {\small\ttfamily <table>} elements
will be treated as values of the same type.

@@

When introducing type providers (\S\ref{sec:providers-sum}), we mentioned that F\# Data 
implements a special handling of heterogeneous collections. This allows us to avoid inferring a 
labelled top shapes in many common scenarios. In the earlier example, a sample collection 
contains a record (with \strf{pages} and \strf{page} fields) and a nested collection with values.

Rather than storing a single shape for the collection elements as in $[\sigma]$, heterogeneous
collections store multiple possible element shapes together with their \emph{  inferred multiplicity} 
(zero or one, exactly one, zero or more):
%
\begin{equation*}
\begin{array}{rcl}
 \psi &\narrow{=}& 1? \lsep 1 \lsep \ast \\
 \sigma &\narrow{=}& ~\ldots \lsep [\sigma_1, \psi_1 | \ldots | \sigma_n, \psi_n ]
\end{array}
\end{equation*}
%
We omit the details, but finding a preferred common shape of two heterogeneous 
collections is analogous to the handling of labelled top types. We merge cases with the same tag (by finding 
their common shape) and calculate their new shared multiplicity (for example, by turning 
$1$ and $1?$ into $1?$).

@@

\paragraph{Extending the type systems.} 
Previous work that integrates external data, such as XML \cite{xduce,xduce-ml} and databases 
\cite{links,linq}, into a programming language, required the user to define the schema
or had an ad-hoc extension that reads the schema.

C$\omega$ \cite{comega-xs} is the most similar to F\# Data. It extends C\# with types similar to our
structural types (including similar heterogeneous collections), but it does not infer the types
from a sample. It also extends the type system of the host language (rather than using a general purpose
embedding).

@@

The PADS project \cite{pads-dsl,pads-ml} tackles a more general problem of handling \emph{any} data format.
The schema definitions in PADS are similar to our structural type. The structure inference for LearnPADS
\cite{pads-learn} infers the data format from a flat input stream. A PADS type provider could follow
many of the patterns we explore in this paper, but formally specifying the safety property would be
more challenging.

@@

We explored the F\# Data type providers for XML, CSV and JSON. As most real-world data do not come 
with an explicit schema, the library uses \emph{shape inference} that deduces a shape from a set of 
samples. Our inference algorithm is based on a preferred shape relation. It prefers records to 
encompass the open world assumption and support developer tooling. The algorithm is predictable, which is 
important as developers need to understand how changing the samples affects the resulting types.

We explore the theory behind type providers. F\# Data is a prime example of 
type providers, but our work demonstrates a more general point. The types generated by type 
providers can depend on external input and so we can only guarantee \emph{relativized safety}, 
which says that a program is safe only if the actual inputs satisfy additional conditions.

Type providers have been described before, but this paper is novel in that it explores the 
properties of type providers that represent the ``types from data'' approach. Our experience suggests 
that this significantly broadens the applicability of statically typed languages to real-world 
problems that are often solved by weakly-typed techniques.
